Illuminating a Dark Lens : A Type Ia Supernova \\ Magnified by the Galaxy Cluster Abell 2744

### 2015-05-17 snTomas.tex
Adjustments to the host galaxy section

\begin{abstract}
\change{Reduced to $<250$ words.}\\
SN \tomas\ is a Type Ia Supernova (SN) discovered at
$z=1.3457\pm0.0001$ behind the galaxy cluster Abell 2744
($z=0.308$). This SN 
%has a projected separation from the cluster core
%of $\sim$40\arcsec, 
is closer to the critical lensing line than any
previous cluster-lensed Type Ia SN, 
%As such, it provides 
providing a valuable
opportunity to test galaxy cluster lens models
with the direct measurement of an absolute magnification at the edge
of the strong-lensing region.  
%We derive a tightly constrained measure
%of the peak apparent magnitude, corrected for light curve shape and
%extinction.  
In a cosmology-independent analysis, we find that \tomas\ is
$0.77\pm0.15$ magnitudes brighter than unlensed Type Ia SNe at similar
redshift, implying a lensing magnification of 
$\mu_{\rm obs}=2.03\pm0.29$. 
\change{In future lens models of Abell 2744, this could
provide some leverage for breaking degeneracies that positional
constraints cannot disentangle.}  We compare the measured $\mu$
against predicted magnifications from 15 lens models, \change{ of
which 13 were generated before the SN magnification was known,
qualifying as a pure ``blind test.''}  The models are collectively
fairly accurate: \change{6 of the blind models and both of the post-SN
models} deliver median magnifications that are consistent with the
measured $\mu$ to within 1$\sigma$.  However, there is a subtle
systematic bias: all of the models that disagree with the SN $\mu$ 
%by more than $1\sigma$ 
are {\it higher} than the measured value and 6 of these
are $\geq1.5\sigma$ above the measured magnification.  We
evaluate possible causes for this mild bias,
%and reject explanations
%that invoke misinterpretation of the SN data or errors in the
%definition of the lens models. We 
and find that
% -- at least for this
%cluster and this line of sight -- 
the most accurate models are those that employ stringent quality cuts
\change{and use a large fraction of spectroscopically measured 
redshifts of multiple images.}
% A more robust evaluation
%\change{Further examination} of systematic biases 
%could be achieved 
%\change{could be pursued with more}
%lens model comparisons using lensed
%Type Ia SNe, as well as cluster lensing simulations.
\end{abstract}

@@

One of the key values in observing standard candles behind
gravitational lenses is in addressing the problem of the
mass-sheet degeneracy \citep{Falco:1985,Schneider:1995}.  This
degeneracy arises because one can introduce into a lens model an
unassociated sheet of uniform mass in front of or behind the
lens, without disturbing the primary observable quantities.  For
example, take a lens model with a given surface mass density $\kappa$,
and then transform the surface mass density to
$\kappa^{\prime}=(1-\lambda)\kappa+\lambda$ for any arbitrary value
$\lambda$.  Both the $\kappa$ and $\kappa^{\prime}$ models will
produce exactly the same values for all positional and shear
constraints from strong and weak lensing \citep{Seitz:1997}.  When lensed background
sources are available across a wide range of redshifts (as is the case
for the Abell 2744 cluster discussed here), it should in principle be
possible to break this degeneracy \citep{Seitz:1997,Bradac:2004}.  However, there are more complex versions
of positional constraint
degeneracies \citep{Schneider:2014,Liesenborgs:2012}.  Such
degeneracies do not extend to the {\it absolute} magnification of a
background source's flux and size. \change{In the case of the simple mass sheet degeneracy described above, the magnification scales as $\mu\propto(1-\lambda)^{-2}$ \citep[see e.g.][]{Bartelmann:2010}.} Therefore, an absolute measurement
of magnification from a standard candle or a standard ruler
\citep{Sonnenfeld:2011}  can break these fundamental degeneracies
\citep{Holz:2001}.

In Section~\ref{sec:DiscoveryAndFollowup} we present the discovery and
follow-up observations of SN \tomas\ at $z=1.3457$, discovered behind
the galaxy cluster Abell 2744.  Sections~\ref{sec:Spectroscopy}
and \ref{sec:PhotometricClassification} describe the spectroscopy and
photometry of this SN, leading to a classification of the object as a
normal Type Ia SN.  In Section~\ref{sec:DistanceAndMagnification} we
make a direct measurement of the magnification of this source due to
gravitational lensing.  Section~\ref{sec:Discussion} discusses the
tension between our magnification measurement and the lens models,
and our conclusions are summarized in Section~\ref{sec:SummaryAndConclusions}.
with implications for the systematic error budget in magnification
estimates for other high-$z$ lensed sources.

@@

@@

@@

It is now an accepted practice in cosmological analyses using Type Ia
SN to apply a correction to the luminosity of each SN based on the
stellar mass of its host galaxy. Typically this is described as a
simple bifurcation of the SN population: SNe that appear in more
massive hosts are observed to be $\sim$0.08 mag brighter (after
corrections for light curve shape and color) than SNe in low-mass
hosts \citep{Kelly:2010,Sullivan:2010}.  The dividing line for this
purely empirical ``mass step'' correction is generally set around
$10^{10}$\Msun, although this is somewhat arbitrary \citep[see
e.g.][]{Betoule:2014}.  This happens to be very close to the
SN \tomas\ host galaxy mass, which we have measured to be
$10^{10.1}$ \Msun\ based on Eq. 7 of \citet{Taylor:2011}, using the
{\it BPZ} SED fit (Section~\ref{sec:DiscoveryAndFollowup}) to define rest-frame optical magnitudes.

@@

@@

The seven free-form models divide the cluster field into a grid,
generally using a multi-scale grid to get better sampling in regions with a higher density of information. Each grid cell is assigned a mass or a potential, and then
the mass values are iteratively refined to match the observed lensing
constraints. In some cases an adaptive grid is used so that the grid
spacing itself can also be modified as the model is
iterated \citep[e.g.][]{Liesenborgs:2006,Merten:2009,Bradac:2009}.
\change{ In general, free-form methods tend to be more flexible than 
parametric methods, as they are not guided (or restricted) by
assumptions about the mass distribution of dark matter halos. }

@@

>>>>

\begin{abstract}
\changenote{Reduced to $<250$ words.}\\
SN \tomas\ is a Type Ia Supernova (SN) discovered at
$z=1.3457\pm0.0001$ behind the galaxy cluster Abell 2744
($z=0.308$). This SN 
%has a projected separation from the cluster core
%of $\sim$40\arcsec, 
is closer to the critical lensing line than any
previous cluster-lensed Type Ia SN, 
%As such, it provides 
providing a valuable
opportunity to test galaxy cluster lens models
with the direct measurement of an absolute magnification at the edge
of the strong-lensing region.  
%We derive a tightly constrained measure
%of the peak apparent magnitude, corrected for light curve shape and
%extinction.  
In a cosmology-independent analysis, we find that \tomas\ is
$0.77\pm0.15$ magnitudes brighter than unlensed Type Ia SNe at similar
redshift, implying a lensing magnification of 
$\mu_{\rm obs}=2.03\pm0.29$. 
\change{In future lens models of Abell 2744, this could
provide some leverage for breaking degeneracies that positional
constraints cannot disentangle.}  We compare the measured $\mu$
against predicted magnifications from 15 lens models, \change{ of
which 13 were generated before the SN magnification was known,
qualifying as a pure ``blind test.''}  The models are collectively
fairly accurate: \change{6 of the blind models and both of the post-SN
models} deliver median magnifications that are consistent with the
measured $\mu$ to within 1$\sigma$.  However, there is a subtle
systematic bias: all of the models that disagree with the SN $\mu$ 
%by more than $1\sigma$ 
are {\it higher} than the measured value and 6 of these
are $\geq1.5\sigma$ above the measured magnification.  We
evaluate possible causes for this mild bias,
%and reject explanations
%that invoke misinterpretation of the SN data or errors in the
%definition of the lens models. We 
and find that
% -- at least for this
%cluster and this line of sight -- 
the most accurate models are those that employ stringent quality cuts
\change{and use a large fraction of spectroscopically measured 
redshifts of multiple images.}
% A more robust evaluation
%\change{Further examination} of systematic biases 
%could be achieved 
%\change{could be pursued with more}
%lens model comparisons using lensed
%Type Ia SNe, as well as cluster lensing simulations.
\end{abstract}

@@

One of the key values in observing standard candles behind
gravitational lenses is in addressing the problem of the mass-sheet
degeneracy \citep{Falco:1985,Schneider:1995}.  This degeneracy arises
because one can introduce into a lens model an unassociated sheet of
uniform mass in front of or behind the lens, without disturbing the
primary observable quantities.  For example, take a lens model with a
given surface mass density $\kappa$, and then transform the surface
mass density to $\kappa^{\prime}=(1-\lambda)\kappa+\lambda$ for any
arbitrary value $\lambda$.  Both the $\kappa$ and $\kappa^{\prime}$
models will produce exactly the same values for all positional and
shear constraints from strong and weak lensing \citep{Seitz:1997}.
When lensed background sources are available across a wide range of
redshifts (as is the case for the Abell 2744 cluster discussed here),
it should in principle be possible to break this
degeneracy \citep{Seitz:1997,Bradac:2004}.  However, there are more
complex versions of positional constraint
degeneracies \citep{Liesenborgs:2012,Schneider:2014}.  Such
degeneracies do not extend to the {\it absolute} magnification of a
background source's flux and size. \change{In the case of the simple
mass sheet degeneracy described above, the magnification scales as
$\mu\propto(1-\lambda)^{-2}$ \citep[see e.g.][]{Bartelmann:2010}.}
Therefore, one can break these fundamental degeneracies with an
absolute measurement of magnification from a standard
candle \citep{Holz:2001} or a standard ruler \citep{Sonnenfeld:2011}.

In Section~\ref{sec:DiscoveryAndFollowup} we present the discovery and
follow-up observations of SN \tomas\ at $z=1.3457$, discovered behind
the galaxy cluster Abell 2744.  Section~\ref{sec:HostGalaxy} examines
the SN host galaxy.  Sections~\ref{sec:Spectroscopy}
and \ref{sec:PhotometricClassification} describe the spectroscopy and
photometry of this SN, leading to a classification of the object as a
normal Type Ia SN.  In Section~\ref{sec:DistanceAndMagnification} we
make a direct measurement of the magnification of this source due to
gravitational lensing.  Section~\ref{sec:Discussion} discusses the
tension between our magnification measurement and the lens models. Our
conclusions are summarized in Section~\ref{sec:SummaryAndConclusions},
along with a discussion of future prospects.

@@

\section{Host Galaxy}
\label{sec:HostGalaxy}

@@

\change{
The galaxy identified as the host is not close enough to the cluster
core to {\it necessarily} be multiply-imaged, but it is still possible
that the host galaxy is one of the outer images of a multiple image
system.  In such a case, and given the position of the SN host, one
would expect that another image of the galaxy would be present at a
similar brightness, and would therefore be detectable in HST
imaging. To date, no plausible candidate for a counter-image has been
identified.}

\change{To measure the stellar mass of the host galaxy
we use Eq. 8 of \citet{Taylor:2011}, which relates the rest-frame
(g-i) color and i-band luminosity to the total stellar mass. To derive
these values, we fixed the redshift at $z=1.3457$ and repeated the SED
fitting using BPZ. From the best-fit SED we extracted rest-frame
optical magnitudes, and corrected them for lensing using a
magnification factor of $\mu=2.0$ -- a value that we will derive from
the SN itself in Section~\ref{sec:DistanceAndMagnification}.  From
this we determine the host galaxy mass to be $10^{10.1}$ \Msun. }

@@

It is now an accepted practice in cosmological analyses using Type Ia
SN to apply a correction to the luminosity of each SN based on the
stellar mass of its host galaxy. Typically this is described as a
simple bifurcation of the SN population: SNe that appear in more
massive hosts are observed to be $\sim$0.08 mag brighter (after
corrections for light curve shape and color) than SNe in low-mass
hosts \citep{Kelly:2010,Sullivan:2010}.  The dividing line for this
purely empirical ``mass step'' correction is generally set around
$10^{10}$\Msun, although this is somewhat arbitrary \citep[see
e.g.][]{Betoule:2014}.

@@

\change{
If we were to apply the correction to this sample, then the SN \tomas\
distance modulus would not be adjusted, because its host galaxy mass
(Section~\ref{sec:HostGalaxy}) is below 10$^{10}$ \Msun. Note that
there is some circularity here, as we used the SN magnification in
deriving the host galaxy mass, and now use the host mass to inform the
magnification.  However, changing the magnification value to $mu=2.5$
does not move the \tomas\ host mass above the mass step threshold, so
the circularity is not relevant in this case. Applying the
mass step correction across the entire control sample would lead to a
small adjustment of the inferred SN \tomas\ magnification, but it is $<5\%$
and therefore could not relieve the observed tension.}

@@

The seven free-form models divide the cluster field into a grid,
generally using a multi-scale grid to get better sampling in regions
with a higher density of information (e.g. density of multiple
images). Each grid cell is assigned a mass or a potential, and then
the mass values are iteratively refined to match the observed lensing
constraints. In some cases an adaptive grid is used so that the grid
spacing itself can also be modified as the model is
iterated \citep[e.g.][]{Liesenborgs:2006,Merten:2009,Bradac:2009}.

\change{
As usual, there is a tradeoff between a model's flexibility, the
strength of the model assumptions, and the resulting uncertainties. In
a probabilistic framework, the posterior distribution function of the
desired quantities depends on all the priors, including model
assumptions like parametrization.  In general, free-form methods tend
to be more flexibile than simply parameterized models, and thus tend
to result in larger error bars.  In brief, if the free form methods
are too flexible, then they will result in overestimated error
bars. Conversely, if the simply parameterized models are too
inflexible, they will result in underestimated error bars.
}

@@

In
addition, some sources of uncertainty are taken into account only by a
subset of models. For example, the Willams model (using GRALE;
Liesenborgs et al. 2006, 2007; Mohammed et al. 2014) accounts for mass
sheet degeneracy by introducing a free parameter in the reconstruction
that represents an arbitrarily scaled mass sheet. The flexibility of
the Williams model and the inclusion of this source of uncertainty
contribute to its relatively large uncertainties compared to the other
models.
